--- 
title: "Multivariate and Categorical Statistics"
author: "Daniela Pinto Veizaga & Sebastián Cadavid Sánchez"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
  
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: https://github.com/C1587S/EMV-ejercicios
description: "This is a book meant to me read along side the book Multivariate Methods."

---

# Prerequisites

Placeholder



<!--chapter:end:index.Rmd-->

# Introduction {#intro}

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].

<!--chapter:end:01-intro.Rmd-->

# Literature

Here is a review of existing methods.


HEllo. 

<!--chapter:end:02-literature.Rmd-->

# Methods

We describe our methods in this chapter.

<!--chapter:end:03-method.Rmd-->


# Test on one or two mean vectors {#chapter5}

Placeholder


## Multivariate Test for $H_{0}:\mu=\mu_{0}$ with $\Sigma$ known
## Multivariate Test for $H_{0}:\mu=\mu_{0}$ with $\Sigma$ known
## Multivariate Two-Sample $T^2$-test
### 5.5 Tests on individual variables conditional on rejection of $H_{0}$ by the $T^2$-test.
## Computation of $T^2$
## Paired Observation Test
## Test for Aditional Information
## Profile Analysis
#### Parallelism
#### Same overall level
#### Flatness

<!--chapter:end:04-chapter5.Rmd-->


# Multivariate Analysis of Variance Vectors {#chapter6}

Placeholder


## One Way Models
##### Measures of Multivariate Association
## Comparison of the four MANOVA test statistics comparison ot the four MANOVA test statistics
## Constrasts
### Univariate Contrasts
### Multivariate Contrasts
## Tests on individual variables following rejection of $H_0$ by the overall MANOVA test
##  Two-way Classification
## Other Models
## Checking on Assumptions
## Profile Analysis
### Repeated Measures Designs
## Growth Curves
## Tests on a Subvector

<!--chapter:end:05-chapter6.Rmd-->


# Tests in covariance matrices {#chapter7}

Placeholder


## Introduction
## Testing a specified pattern for $\Sigma$
### Testing $H_0: \Sigma=\Sigma_0$
### Testing Sphericity
### Testing $H_0:\Sigma=\sigma^2[(1-\rho)I+\rho J]$
## Tests comparing covariance matrices
### Univariate Tests of Equality of Variances
### Multivariate Tests of Equality of Covariance Matrices
## Tests of independence
### Independence of Two Subvectors
### Independence of Several Subvectors
### Test for Independence of All Variables

<!--chapter:end:06-chapter7.Rmd-->


# Discriminant Analysis: description of group separation {#chapter8} 

Placeholder


## Introduction
## Discriminant Function for Two Vectors
### Example 8.2
## Relantionship Between Two Group Discriminant Analysis and Multiple Regression
### Example 8.3.
## Discriminant Analysis For Several Groups
### Example 8.4.1.
### Example 8.4.2.
## Standarized Discriminant Functions
## Test of Significance
### Example 8.6.2
## Interpretation of Discriminant Functions
## Scatter Plots
### Example 8.8.
## Stepwise Selection of Variables
### Example 8.9.

<!--chapter:end:07-chapter8.Rmd-->


# Classification Analysis: Allocation of Observartion Groups {#chapter9}

Placeholder


## Introduction
## Classification into Two Groups
### Example 9.2.
## Classification into Several Groups
### Example 9.3
## Estimating Misclassification Rates
### Example $9.4.(a).$ 
### Example $9.4.(b)$ 
## Improved Estimates of Error Rates
### Example $9.5.2$
## Subset Selection
### Example 9.6.(a).
### Example 9.6.(b).
## Nonparametric Procedures
### Example 9.7.2.

<!--chapter:end:08-chapter9.Rmd-->

# Canonical Correlation {#chapter11}


## Introduction

## Canonical Correlations and Canonical Variates

## Properties of Canonical Correlations

### Example 11.3

For the chemical data of Table $10.1$ we obtain the canonical correlations and illustrate property $2$. We consider the extended set of nine $x's$, as in Example $10.5.2$. The matrix $R_{yx}$ of correlations between the $y's$ and the $x's$ is:

```{r}
experiment_number <- seq(1:19)

y1 <- c(
  41.5, 33.8, 27.7, 21.7, 19.9, 15, 12.2, 4.3, 19.3, 6.4, 37.6, 18, 26.3,9.9, 25, 14.1, 15.2, 15.9, 19.6
  )
  
y2 <- c(
  45.9, 53.3, 57.5, 58.8, 60.6, 58, 58.6, 52.4, 56.9, 55.4,46.9, 57.3, 55, 58.9, 50.3, 61.1, 62.9, 60, 60.6
  )

y3 <- c(
  11.2, 11.2, 12.7, 16, 16.2, 22.6, 24.5, 38, 21.3, 30.8, 14.7, 22.2, 18.3, 28, 22.1, 23, 20.7,22.1, 19.3
)

x1 <- c(
  162, 162, 162, 162, 172, 172, 172, 172, 167, 177, 157, 167, 167, 167, 167,177,177, 160, 160
)

x2 <- c(
  23, 23, 30, 30, 25, 25, 30, 30, 27.5, 27.5, 27.5, 32.5, 22.5, 27.5, 27.5, 20, 20, 34, 34
)

x3 <- c(
  3, 8, 5, 8, 5, 8,5, 8, 6.5, 6.5, 6.5, 6.5, 6.5, 9.5, 3.5, 6.5, 6.5, 7.5, 7.5
)

x1x1 <- x1*x1
x2x2 <- x2*x2
x3x3 <- x3*x3
x1x2 <- x1*x2
x1x3 <- x1*x3
x2x3 <- x2*x3
chemicalReaction <- data.frame(experiment_number, y1, y2, y3, x1, x2, x3, x1x1, x2x2, x3x3, x1x2, x1x3, x2x3)
chemicalReaction
```


```{r tidy=FALSE}
knitr::kable(
  chemicalReaction, caption = 'Chemical Reaction Data',
  booktabs = TRUE
)
```



```{r}
if(!require("magrittr")){
    install.packages("magrittr")
    library(magrittr)
}


Y<-chemicalReaction%>% select(2:4)
X <- chemicalReaction%>% select(5:13)
X
```



Printing out the $R_{xx}$, $R_{yy}$, $R_{xy}$ and $R_{yx}$ in that order:


```{r}

if(!require("CCA")){
    install.packages("CCA")
    library(CCA)
}

correl<-matcor(X, Y)
correl


```

Plotting the correl:
```{r}
img.matcor(correl, type = 2)
```


The three canonical correlations are obtain via the `cancor` package in R.

```{r}
stats::cancor(X, Y)
```



Therefore, the three canonical correlations are:

$$$$

$$$$

$$$$



## Tests of Significance

### Example 11.4.1.

For the chemical data of Table $10.1$ with the extended set of nine $x's$, we obtained canonical correlations 
### Example 11.4.2.

## Interpretation

### Example 11.5.1.

For the chemical data in Table $10.1$ with the extended set of nine $x's$, we obtain the following standarized coefficients for the three canonical variates:



## Relationships of Canonical Correlation Analysis to Other Multivariate Techniques

## References


<!--chapter:end:09-chapter11.Rmd-->

